---
title: "MA Fitness"
author: "Rutuja Gupte"
date: "2023-09-21"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "MA Fitness"
author: "Dr Nathaniel Sharp"
---

- this fits a spline to y=OD versus x=time, and then finds the maximum slope
```{r}
spline.slope<-function(x, y, n=101, eps=1e-5, span=0.2){
  max(nderiv(loess(log(y) ~ x, degree=1, span=span), seq(min(x), max(x), length=n)), na.rm=TRUE)
}
```

- used by the function above to get a local (linear) slope around a point

```{r}
nderiv <- function(fit, x, eps=1e-5){
  (predict(fit, x + eps) - predict(fit, x - eps))/(2 * eps)}

```

---
title: "MA Fitness"
author: "Rutuja Gupte"
---

Import Statements
```{r}
library(dplyr)
library(stringr)
library(tidyverse)
library(lme4)
library(vioplot)
```

Load all files and setup variables
```{r}
l <- list.files()

# read all data into a list of data frames
dfs <- lapply(l[endsWith(l, '.csv')], function(r){
  d <- read.csv(r)
  
  assay.name <- str_replace(str_replace(r, 'Ref', 'Rutuja'), '.csv', '.txt')
  assay.data <- read.delim(assay.name)
  
  # Change time column to 0, 0.25, 0.5, ...
  assay.data$Time<-seq(from=0.25,by=0.25,length.out=nrow(assay.data))
  
  # calculate slopes
  d$slope <- sapply(c(1:nrow(d)), function(r){
    spline.slope(assay.data$Time, assay.data[,which(names(assay.data)==d$well[r])])
  })
  
  d$initial <- sapply(c(1:nrow(d)), function(r){
    mean(assay.data[,which(names(assay.data)==d$well[r])][1:4])
  })
  
  d$final <- sapply(c(1:nrow(d)), function(r){
    temp <- assay.data[,which(names(assay.data)==d$well[r])]
    temp <- temp[!is.na(temp)]
     n <- length(temp)
    mean(assay.data[,which(names(assay.data)==d$well[r])][n-3:n])
  })

  d$date <- mdy(str_replace(str_replace(r, 'Ref ', ''), '.csv', ' 2023'))
  return(d)
})

# merge the list of data frames
d <- dfs %>% reduce(bind_rows)

dates.1 <- seq(mdy(04062023), mdy(04102023), 1)
dates.2 <- seq(mdy(04282023), mdy(05052023), 1)
# dates.2 <- c(dates.2, mdy(11022023))

d <- d %>% 
  filter(date %in% dates.2)

ancestors <- c('H1', 'H2', 'H3', 'D1', 'D2', 'D3', 'C1', 'C3')
# these are all the haploid ancestors including the diploids that are supposed to be diploids
ancestors.haploid <- c('H1', 'H2', 'H3', 'D2', 'D3', 'C1', 'C3')
# these are haploids that were intended to be haploids
anc.hap.og <- c('H1', 'H2', 'H3')
# these are haploids that were intended to be diploids
anc.hap.fake <- c('D2', 'D3')
# this is the only actual diploid control that was diploid
ancestors.diploid <- c('D1')
# this MA line started out as a haploid but ended up as a diploid
fake.haploids <- c(81)
# these MA lines started out as diploids but ended up as haploids
fake.diploids <- c(20, 28, 48, 52, 54, 84, 100)
```

Adding additional information for categorization. Exluding MA lines that changed ploidy. 
```{r}
d$label <- d$treatment
d$label[d$treatment == 'Blank'] <- '0'
d$label[d$treatment == 'H1'] <- '101'
d$label[d$treatment == 'D1'] <- '102'
d$label[d$treatment == 'H2'] <- '103'
d$label[d$treatment == 'D2'] <- '104'
d$label[d$treatment == 'H3'] <- '105'
d$label[d$treatment == 'D3'] <- '106'
d$label[d$treatment == 'C1'] <- '107'
d$label[d$treatment == 'C3'] <- '109'
d$label <- as.numeric(d$label)

d <- d %>% mutate(category = case_when(label == 0 ~ 'Blank',
                              label > 100 & treatment %in% ancestors.haploid ~ 'Ctrl.H',
                              label > 100 & treatment %in% ancestors.diploid ~ 'Ctrl.D',
                              label %% 2 == 0 ~ 'MA.D',
                              label %% 2 == 1 ~ 'MA.H'))

d <- d %>% mutate(category = ifelse(label %in% fake.diploids, 'MA.H', category))
d <- d %>% mutate(category = ifelse(label %in% fake.haploids, 'MA.D', category))
```


> Outlier removal

Make predictions for the blanks. For that, remove the blank outliers that seem to be contaminated.
```{r}
blanks <-  d %>% filter(category == 'Blank')
blanks %>% ggplot() + geom_point(aes(x=slope, y=1)) +
  scale_x_log10()

# setting cutoff for blanks around 
good.blanks <- blanks %>% filter(slope < 0.01)
bad.blanks <- blanks %>% filter(slope > 0.01)

model <- lmer(slope~(1|date), data = good.blanks)
dates.predict <- data.frame(date=distinct(d, date))
dates.predict$null = predict(model, dates.predict)
```

Remove experimental errors using rules:
1. Remove any slopes less than the predicted blank slopes.
```{r}
df <- anti_join(d, bad.blanks)
df <- full_join(df, dates.predict, by=join_by(date))
df <- df %>% filter(slope >= null)
```

2. Remove any curves where final OD was less than initial OD.
```{r}
df$delta <- df$final - df$initial
df <- df %>% filter(delta >= 0)
```

3. Remove ancestors that changed ploidy
```{r}
df <- df %>% 
  filter(!(treatment %in% anc.hap.fake))
```


> Data Analysis

```{r}
data <- df %>% select(date, well, treatment, label, category, slope, initial) %>%
  filter(category != 'Blank')

data$ploidy <- case_when(
  data$category %in% c('Ctrl.H', 'MA.H') ~ 'Haploid',
  data$category %in% c('Ctrl.D', 'MA.D') ~ 'Diploid'
)

data$MA <- case_when(
  data$category %in% c('MA.H', 'MA.D') ~ 'MA',
  data$category %in% c('Ctrl.H', 'Ctrl.D') ~ 'Ctrl',
  data$category == 'Blank' ~ 'Blank'
)

ctrl <- data %>% filter(label > 100)
trt <- data %>% filter(label > 0 & label <= 100)
```

Calculating the fitness of ancestors for each day
```{r}
mod <- lmer(slope ~ ploidy + (1|date), ctrl)
ctrl.predict <- data.frame(distinct(ctrl, date, ploidy))
ctrl.predict$ctrl <- predict(mod, ctrl.predict)
```

Calculating relative fitness
```{r}
trt <- left_join(trt, ctrl.predict, by=c('date', 'ploidy'))
trt$rel.fit <- trt$slope - trt$ctrl

trt <- trt %>% group_by(ploidy, label) %>%
  summarize(rel.fit = mean(rel.fit)) %>%
  ungroup()
```

Computing summary statistics
```{r}
mu.dip <- mean(trt$rel.fit[trt$ploidy == 'Diploid'])
mu.dip
sd.dip <- sd(trt$rel.fit[trt$ploidy == 'Diploid'])
sd.dip

mu.hap <- mean(trt$rel.fit[trt$ploidy == 'Haploid'])
mu.hap
sd.hap <- sd(trt$rel.fit[trt$ploidy == 'Haploid'])
sd.hap
```

Plotting relative fitness
```{r}
trt %>% ggplot() +
  geom_histogram(aes(x=rel.fit), fill='black') +
  facet_grid(rows = vars(ploidy)) +
  geom_vline(data = data.frame(ploidy = c('Haploid', 'Diploid'),
                               mu = c(mu.hap, mu.dip)),
             aes(xintercept = mu, color = ploidy)) +
  geom_vline(xintercept = 0, color = 'black') +
  xlab("Relative Fitness") +
  ylab("Number of Lines") +
  guides(color = guide_legend(title = "Ploidy"))
```

Violin plot of relative fitness
```{r}
vioplot(trt[trt$ploidy == 'Haploid',]$rel.fit,
        trt[trt$ploidy == 'Diploid',]$rel.fit,
        names = c("Haploid", "Diploid"),
        col = c('red', 'blue'),
        ylim = c(-0.15, 0.01), 
        ylab = "Relative Fitness",
        xlab = "Ploidy")
abline(h=0)
```


Using lmer for hypothesis testing
```{r}
null <- lmer(slope ~ MA + ploidy + initial + (1|date) + (1|treatment), data)
summary(null)
full <- lmer(slope ~ MA*ploidy + initial + (1|date) + (1|treatment), data)
summary(full)

mod <- anova(null, full)
mod
```

Combining with MA data
```{r}
mut <- read_delim("pombe_MA_data.txt")
mut <- mut %>% rename('ploidy.ma' = 'ploidy')
trt <- trt %>%
  left_join(mut, by=c('label'='line'))

trt <- trt %>% select(ploidy, label, rel.fit, ploidy.ma, ploidy_final,
                        n.SNM, n.indel)
trt <- trt %>% mutate(mutations = n.SNM + n.indel)
```

Plot for relation between mutations and relative fitness
```{r}
trt %>% ggplot() +
  geom_point(aes(x=mutations , y=rel.fit, color=ploidy)) +
  geom_smooth(data=trt[trt$ploidy == 'Haploid',],
              mapping=aes(x=mutations, y=rel.fit), color='Blue', method="lm", se=FALSE) +
  geom_smooth(data=trt[trt$ploidy == 'Diploid',],
              mapping=aes(x=mutations, y=rel.fit), color='Red', method="lm", se=FALSE) +
  ylab("Relative Fitness") +
  xlab("Number of Mutations (SNM + indel)") +
  guides(color = guide_legend(title = "Ploidy"))
```

```{r}
mod <- lm(rel.fit ~ mutations, trt)
summary(mod)

mod <- lm(rel.fit ~ mutations, trt[trt$ploidy == 'Haploid',])
summary(mod)

mod <- lm(rel.fit ~ mutations, trt[trt$ploidy == 'Diploid',])
summary(mod)
```


To create the data appendix
```{r}
write.csv(d, "../data_ip_draft1.csv")
```



