---
title: "MA Fitness"
author: "Rutuja Gupte"
date: "2023-09-21"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "MA Fitness"
author: "Dr Nathaniel Sharp"
---

- this fits a spline to y=OD versus x=time, and then finds the maximum slope
```{r}
spline.slope<-function(x, y, n=101, eps=1e-5, span=0.2){
  max(nderiv(loess(log(y) ~ x, degree=1, span=span), seq(min(x), max(x), length=n)), na.rm=TRUE)
}
```

- used by the function above to get a local (linear) slope around a point

```{r}
nderiv <- function(fit, x, eps=1e-5){
  (predict(fit, x + eps) - predict(fit, x - eps))/(2 * eps)}

```

---
title: "MA Fitness"
author: "Rutuja Gupte"
---

Import Statements
```{r}
library(dplyr)
library(stringr)
library(tidyverse)
library(lme4)
```

> Setting up functions to extract information from the original data.

- this fits a spline to y=OD versus x=time, and then finds the time-stamp for maximum slope
```{r}
spline.time<-function(x, y, n=101, eps=1e-5, span=0.2){
  which.max(nderiv(loess(log(y) ~ x, degree=1, span=span), seq(min(x), max(x), length=n)))
}
```

- this fits a linear regression model and extracts the slope
```{r}
well.lm.slope <- function(x, y){
  lm(y ~ x)$coefficients[2]
}
```

Load all files and setup variables
```{r}
l <- list.files()

# read all data into a list of data frames
dfs <- lapply(l[endsWith(l, '.csv')], function(r){
  d <- read.csv(r)
  
  assay.name <- str_replace(str_replace(r, 'Ref', 'Rutuja'), '.csv', '.txt')
  assay.data <- read.delim(assay.name)
  # Change time column to 0, 0.25, 0.5, ...
  assay.data$Time<-seq(from=0.25,by=0.25,length.out=nrow(assay.data))
  # calculate slopes
  d$slope <- sapply(c(1:nrow(d)), function(r){
    spline.slope(assay.data$Time, assay.data[,which(names(assay.data)==d$well[r])])
  })
  
  d$initial <- sapply(c(1:nrow(d)), function(r){
    mean(assay.data[,which(names(assay.data)==d$well[r])][1:4])
  })
  
  d$final <- sapply(c(1:nrow(d)), function(r){
    temp <- assay.data[,which(names(assay.data)==d$well[r])]
    temp <- temp[!is.na(temp)]
     n <- length(temp)
    mean(assay.data[,which(names(assay.data)==d$well[r])][n-3:n])
  })

  
  # calculate time stamps for max slope
  d$time <- sapply(c(1:nrow(d)), function(r){
    spline.time(assay.data$Time, assay.data[,which(names(assay.data)==d$well[r])])
  })
  
  #calculate slope of linear model
  d$lm.slope <- sapply(c(1:nrow(d)), function(r){
    well.lm.slope(assay.data$Time, assay.data[,which(names(assay.data)==d$well[r])])
  })
  
  d$time <- assay.data$Time[d$time]
  
  d$date <- mdy(str_replace(str_replace(r, 'Ref ', ''), '.csv', ' 2023'))
  return(d)
})

# merge the list of data frames
d <- dfs %>% reduce(bind_rows)

d <- d %>% 
  filter(!(date %in% seq(mdy(04062023), mdy(04102023), 1)))

ancestors <- c('H1', 'H2', 'H3', 'D1', 'D2', 'D3')
# these are all the haploid ancestors including the diploids that are supposed to be diploids
ancestors.haploid <- c('H1', 'H2', 'H3', 'D2', 'D3')
# these are haploids that were intended to be haploids
anc.hap.og <- c('H1', 'H2', 'H3')
# these are haploids that were intended to be diploids
anc.hap.fake <- c('D2', 'D3')
# this is the only actual diploid control that was diploid
ancestors.diploid <- c('D1')
# this MA line started out as a haploid but ended up as a diploid
fake.haploids <- c(81)
# these MA lines started out as diploids but ended up as haploids
fake.diploids <- c(20, 28, 48, 52, 54, 84, 100)
```

Adding additional information for categorization. Exluding MA lines that changed ploidy. 
```{r}
d <- d %>% mutate(label = recode(treatment,
                                 'Blank' = '0',
                                 'H1' = '101',
                                 'D1' = '102',
                                 'H2' = '103',
                                 'D2' = '104',
                                 'H3' = '105',
                                 'D3' = '106'))
d$label <- as.numeric(d$label)

d <- d %>% mutate(category = case_when(label == 0 ~ 'Blank',
                              label > 100 & treatment %in% ancestors.haploid ~ 'Ctrl.H',
                              label > 100 & treatment %in% ancestors.diploid ~ 'Ctrl.D',
                              label %% 2 == 0 ~ 'MA.D',
                              label %% 2 == 1 ~ 'MA.H'))

# exclude the haploid and diploid MA lines that switched ploidy for now
d <- d %>% filter(!(label %in% fake.diploids)) %>%
  filter(!(label %in% fake.haploids))

```


> Outlier removal

Make predictions for the blanks. For that, remove the blank outliers that seem to be contaminated.
```{r}
blanks <-  d %>% filter(category == 'Blank')
blanks %>% ggplot() + geom_point(aes(x=slope, y=1)) +
  scale_x_log10()

# setting cutoff for blanks around 
good.blanks <- blanks %>% filter(slope < 0.05)
bad.blanks <- blanks %>% filter(slope > 0.05)

model <- lmer(slope~(1|date), data = good.blanks)
dates.predict <- data.frame(date=distinct(d, date))
# dates.predict$batch <- c(rep(1,4), rep(2,4))
dates.predict$null = predict(model, dates.predict)
```

Remove outlier using rules:
1. Remove any slopes less than the predicted blank slopes.
```{r}
df <- anti_join(d, bad.blanks)
df <- full_join(df, dates.predict, by=join_by(date))
df <- df %>% filter(slope >= null)
```

2. Remove any where slopes predicted by linear model were negative.
```{r}
df <- df %>%
  filter(lm.slope >= 0)
```

3. Remove any curves where final OD was less than initial OD.
```{r}
df$delta <- df$final - df$initial
df <- df %>% filter(delta >= 0)
```

4. Remove any curves where the maximum slope was attained in the first hour
```{r}
df <- df %>% filter(time > 1)
```

> Data Analysis

```{r}
data <- df %>% select(date, well, treatment, label, category, slope) %>%
  filter(category != 'Blank') %>%
  filter(!(treatment %in% anc.hap.fake))
data$ploidy <- case_when(
  data$category %in% c('Ctrl.H', 'MA.H') ~ 'Haploid',
  data$category %in% c('Ctrl.D', 'MA.D') ~ 'Diploid'
)
data$MA <- case_when(
  data$category %in% c('MA.H', 'MA.D') ~ 'MA',
  data$category %in% c('Ctrl.H', 'Ctrl.D') ~ 'Ctrl',
  data$category == 'Blank' ~ 'Blank'
)
ctrl <- data %>% filter(label > 100)
trt <- data %>% filter(label > 0 & label <= 100)
```

Calculating the fitness of ancestors for each day
```{r}
mod <- lmer(slope ~ ploidy + (1|date), ctrl)
ctrl.predict <- data.frame(distinct(ctrl, date, ploidy))
ctrl.predict$ctrl <- predict(mod, ctrl.predict)
```

Calculating relative fitness
```{r}
trt <- left_join(trt, ctrl.predict, by=c('date', 'ploidy'))
trt$rel.fit <- trt$slope - trt$ctrl

trt <- trt %>% group_by(ploidy, label) %>%
  summarize(rel.fit = mean(rel.fit)) %>%
  ungroup()
```

Computing summary statistics
```{r}
mu.dip <- mean(trt$rel.fit[trt$ploidy == 'Diploid'])
mu.dip
sd.dip <- sd(trt$rel.fit[trt$ploidy == 'Diploid'])
sd.dip

mu.hap <- mean(trt$rel.fit[trt$ploidy == 'Haploid'])
mu.hap
sd.hap <- sd(trt$rel.fit[trt$ploidy == 'Haploid'])
sd.hap
```

t tests
```{r}
t.test(trt$rel.fit[trt$ploidy == 'Diploid'])
t.test(trt$rel.fit[trt$ploidy == 'Haploid'])

t.test(trt$rel.fit[trt$ploidy == 'Diploid'], 
       trt$rel.fit[trt$ploidy == 'Haploid'],
       var.equal = FALSE)
```

Plotting relative fitness
```{r}
trt %>% ggplot() +
  geom_histogram(aes(x=rel.fit), fill='black') +
  facet_grid(rows = vars(ploidy)) +
  geom_vline(data = data.frame(ploidy = c('Haploid', 'Diploid'),
                               mu = c(mu.hap, mu.dip)),
             aes(xintercept = mu, color = ploidy)) +
  geom_vline(xintercept = 0, color = 'black') +
  xlab("Relative Fitness") +
  ylab("Number of Lines") +
  guides(color = guide_legend(title = "Ploidy"))
```

Using lmer for hypothesis testing
```{r}
null <- lmer(slope ~ MA + ploidy + (1|date) + (1|treatment), data)
summary(null)
full <- lmer(slope ~ MA*ploidy + (1|date) + (1|treatment), data)
summary(full)

mod <- anova(null, full)
mod
```


```{r}
write.csv(d, "data_ip_draft1.csv")
```



