---
title: "MA Fitness v3.Rmd"
author: "Rutuja"
date: "2023-12-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup functions

```{r, include=FALSE}
library(dplyr)
library(stringr)
library(tidyverse)
library(lme4)
library(vioplot)
```

---
title: "MA Fitness"
author: "Dr Nathaniel Sharp"
---

- this fits a spline to y=OD versus x=time, and then finds the maximum slope
```{r}
spline.slope<-function(x, y, n=101, eps=1e-5, span=0.075){
  max(nderiv(loess(log(y) ~ x, degree=1, span=span), x), na.rm=TRUE)
}
```

- used by the function above to get a local (linear) slope around a point

```{r}
nderiv <- function(fit, x, eps=1e-5){
  (predict(fit, x + eps) - predict(fit, x - eps))/(2 * eps)}
```


---
title: "MA Fitness"
author: "Rutuja Gupte"
---

### Spline fitting:

First we start by converting everything to log scale. During the exponential growth phase, log(N) is proportional to t where N is the population and t is time. There is noise in the original data. We remove the noise and try to obtain the underlying curve using loess. It gives a smooth curve. Then we use that curve to find local slopes. nderiv gives precise estimate of the local linear slope at every point.  

- this fits a spline to y=OD versus x=time, and then finds the time-stamp for maximum slope 

```{r}
spline.time<-function(x, y, n=101, eps=1e-5, span=0.075){
  estimates <- loess(log(y) ~ x, degree=1, span=span)
  slopes <- nderiv(estimates, x)
  return(which.max(slopes))
}
```

**Note that the unit of the time stamp is the number of 15 minute intervals from the beginning of measurement.**

## Reading and cleaning the data

Load all files and setup variables
```{r, include=FALSE}
l <- list.files()

# read all data into a list of data frames
dfs <- lapply(l[endsWith(l, '.csv')], function(r){
  d <- read.csv(r)
  
  assay.name <- str_replace(str_replace(r, 'Ref', 'Rutuja'), '.csv', '.txt')
  assay.data <- read.delim(assay.name)
  
  # Change time column to 0, 0.25, 0.5, ...
  assay.data$Time<-seq(from=0.25,by=0.25,length.out=nrow(assay.data))
  
  # calculate the maximum slope of fitted spline
  # maximum growth rate
  d$slope <- sapply(c(1:nrow(d)), function(r){
    spline.slope(assay.data$Time, assay.data[,which(names(assay.data)==d$well[r])], span=0.05)
  })
  
  # initial value OD (in the first hour)
  d$initial <- sapply(c(1:nrow(d)), function(r){
    mean(assay.data[,which(names(assay.data)==d$well[r])][1:4])
  })
  
  # final value of OD (in the last hour when the data was measured)
  # should be a saturated solution
  d$final <- sapply(c(1:nrow(d)), function(r){
    temp <- assay.data[,which(names(assay.data)==d$well[r])]
    temp <- temp[!is.na(temp)]
     n <- length(temp)
    mean(assay.data[,which(names(assay.data)==d$well[r])][n-3:n])
  })
  
  # find the time stamps for the maximum growth rate
  d$time <- sapply(c(1:nrow(d)), function(r){
    spline.time(assay.data$Time, assay.data[,which(names(assay.data)==d$well[r])], span=0.05)
  })
  
  d$date <- mdy(str_replace(str_replace(r, 'Ref ', ''), '.csv', ' 2023'))
  
  d$monotone <- sapply(c(1:nrow(d)), function(r){
    y <- assay.data[,which(names(assay.data)==d$well[r])]
    x <- assay.data$Time
    return(sum(nderiv(loess(log(y) ~ x, degree=1, span=0.2), x) < 0, na.rm=TRUE))
  })
  return(d)
})
```


```{r}
# merge the list of data frames
d <- dfs %>% reduce(bind_rows)

dates.1 <- seq(mdy(04062023), mdy(04102023), 1)
dates.2 <- seq(mdy(04282023), mdy(11302023), 1)
dates.2 <- c(dates.2, mdy(11022023))
ancestors <- c('H1', 'H2', 'H3', 'D1', 'D2', 'D3')

# these are all the haploid ancestors including the diploids that are supposed to be diploids
ancestors.haploid <- c('H1', 'H2', 'H3', 'D2', 'D3', 'C1', 'C3')

# these are haploids that were intended to be haploids
anc.hap.og <- c('H1', 'H2', 'H3', 'C1', 'C2')

# these are haploids that were intended to be diploids
anc.hap.fake <- c('D2', 'D3')

# this is the only actual diploid control that was diploid
ancestors.diploid <- c('D1')

# this MA line started out as a haploid but ended up as a diploid
fake.haploids <- c(81)

# these MA lines started out as diploids but ended up as haploids
fake.diploids <- c(20, 28, 48, 52, 54, 84, 100)

# these have an aneuploidy of chromosome 3
aneuploids <- c(45, 68, 98)
```

Adding additional information for categorization. Excluding MA lines that changed ploidy. Excluding batch 1 due to lack of diploid controls.
```{r}
d$batch[d$date %in% dates.1] <- 1
d$batch[d$date %in% dates.2] <- 2

d <- d %>% filter(batch == 2)

d$label <- d$treatment
d$label[d$treatment == 'Blank'] <- '0'
d$label[d$treatment == 'H1'] <- '101'
d$label[d$treatment == 'D1'] <- '102'
d$label[d$treatment == 'H2'] <- '103'
d$label[d$treatment == 'D2'] <- '104'
d$label[d$treatment == 'H3'] <- '105'
d$label[d$treatment == 'D3'] <- '106'
d$label[d$treatment == 'C1'] <- '107'
d$label[d$treatment == 'C3'] <- '109'
d$label <- as.numeric(d$label)

d <- d %>% mutate(category = case_when(label == 0 ~ 'Blank',
                              label > 100 & treatment %in% ancestors.haploid ~ 'Ctrl.H',
                              label > 100 & treatment %in% ancestors.diploid ~ 'Ctrl.D',
                              label %% 2 == 0 ~ 'MA.D',
                              label %% 2 == 1 ~ 'MA.H'))


d <- d %>% mutate(category = ifelse(label %in% fake.diploids, 'MA.H', category))
d <- d %>% mutate(category = ifelse(label %in% fake.haploids, 'MA.D', category))

# Labeling the dates
dates <- d %>% distinct(date)
alphabet <- c('A', 'B', 'C', 'D', 'E')
dates <- dates %>% mutate(day = alphabet)
d <- d %>% left_join(dates, by='date') %>% select(-date)
```

### Additional information about the days

An additional assay was conducted that compared only the control lines with frozen lab stocks. The ancestors used for the experiment were replicates of the frozen lab stocks. However, they demonstrated considerable variance from the lab stocks. This was a cause for concern which led to the conduction of the supplemental assay. The supplemental assay demonstrated that the control lines used for the original assay behaved similar to the lab stocks. Thus it is reasonable to add these values to the dataset before trimming for reasonable values.  


# Preliminary exploration

Plot ancestors across all days to visually check for day effects
```{r, echo=FALSE}
d %>%
  filter(category == 'Ctrl.H' | category == 'Ctrl.D') %>% ggplot() + geom_point(aes(x=treatment, y=slope, color=treatment)) + facet_wrap(vars(day))

```

Plotting the distribution of the slope values
```{r}
# ci <- c(quantile(d$slope, 0.25) - 1.5* IQR(d$slope),
#         quantile(d$slope, 0.75) + 1.5* IQR(d$slope))

ci <- c(0.05, 0.22)

d %>% ggplot() +
  geom_density(aes(x=slope)) +
  geom_vline(xintercept = ci, color='red', linetype='dashed')
```

Defining the blanks by using a linear model to set a cutoff for the blanks which can then be used to remove the other outliers by predicting the blank values for every day.
```{r}
blanks <-  d %>% filter(category == 'Blank')
```


visualize the blanks
```{r, echo=FALSE}
blanks %>% ggplot() + geom_histogram(aes(x=slope)) +
  scale_x_log10()
```


Setting cutoff for blanks around 0.05 and using the good blanks to make blank predictions for each day.
```{r}
bad.blanks <- blanks %>% filter(slope > 0.05)
good.blanks <- blanks %>% filter(slope < 0.05)

model <- lmer(slope~(1|day), data = good.blanks)
dates.predict <- data.frame(date=distinct(d, day))
dates.predict$null = predict(model, dates.predict)
```


Checking for effects of time (time taken to attain maximum growth rate). Each time stamp is of 15 minutes.

Plotting the low values of time
```{r, echo=FALSE}
# plotting the extremely low values of time
d %>% 
  filter(time<30) %>%
  ggplot() +
  geom_histogram(aes(x=time))
```

Samples with time < 20 were identified to be erroneous. 20 timestamps (4 hours) was a reasonable cutoff since the samples beyond that were more likely to be good.


## Error removal

Accounting for experimental error by removing lines according to the following rules:  
1. Remove unreasonable slope values  
2. Remove the bad blanks.   
3. Remove lines where the initial optical density was less than the final optical   density since the optical density should not decrease unless there was an error.  
4. Remove the diploid ancestors that were found to be haploids.  
5. Remove the lines that reach saturation within the first 4 hours which is too soon to reach saturation.  
```{r}
data <- d %>% 
  filter(slope > ci[1] & slope < ci[2]) %>%
  anti_join(bad.blanks) %>%
  filter(initial <= final) %>%
  filter(!(treatment %in% anc.hap.fake)) %>%
  filter(time > 20)
```

Plotting the ancestors again
```{r, echo=FALSE}
data %>%
  filter(category == 'Ctrl.H' | category == 'Ctrl.D') %>% ggplot() + geom_point(aes(x=treatment, y=slope, color=treatment)) + facet_grid(cols=vars(day))

```

## Preparing for data analysis

```{r}
df <- data %>%
  select(treatment, label, slope, initial, day, category, time) %>%
  filter(category != 'Blank') 

df <- df %>% rename('lineid' = 'treatment')

df$ploidy <- case_when(
  df$category %in% c('Ctrl.H', 'MA.H') ~ 'Haploid',
  df$category %in% c('Ctrl.D', 'MA.D') ~ 'Diploid'
)

df$MA <- case_when(
  df$category %in% c('MA.H', 'MA.D') ~ 'MA',
  df$category %in% c('Ctrl.H', 'Ctrl.D') ~ 'Ctrl',
)

```


# Data Analysis

### Calculating relative fitness

```{r}
# MA lines
trt <- df %>% filter(label > 0 & label <= 100)
# ancestor lines
ctrl <- df %>% filter(label > 100)

# model to predict the slope of the ancestors of each ploidy with date as a random effect
mod <- lmer(slope ~ ploidy + (1|day), ctrl)

# making predictions for each date
ctrl.predict <- data.frame(distinct(ctrl, day, ploidy))
ctrl.predict$ctrl <- predict(mod, ctrl.predict)

# combining with the MA line dateset
trt <- left_join(trt, ctrl.predict, by=c('day', 'ploidy'))

# calculating relative fitness as the difference between the slope for the MA line and its ancestor
trt$rel.fit <- trt$slope - trt$ctrl

# grouping all the readings for the lines and summarizing by calculating the mean relative fitness
trt <- trt %>% group_by(ploidy, label) %>%
  summarize(rel.fit = mean(rel.fit)) %>%
  ungroup()

# mean and standard deviation for the diploids
mu.dip <- mean(trt$rel.fit[trt$ploidy == 'Diploid'])
sd.dip <- sd(trt$rel.fit[trt$ploidy == 'Diploid'])

# mean and standard deviation for the haploids 
mu.hap <- mean(trt$rel.fit[trt$ploidy == 'Haploid'])
sd.hap <- sd(trt$rel.fit[trt$ploidy == 'Haploid'])
```

```{r, echo=FALSE}
trt %>% ggplot() +
  geom_histogram(aes(x=rel.fit), fill='black') +
  facet_grid(rows = vars(ploidy)) +
  geom_vline(data = data.frame(ploidy = c('Haploid', 'Diploid'),
                               mu = c(mu.hap, mu.dip)),
             aes(xintercept = mu, color = ploidy)) +
  geom_vline(xintercept = 0, color = 'black') +
  xlab("Relative Fitness") +
  ylab("Number of Lines") +
  guides(color = guide_legend(title = "Ploidy"))
```

Black line is the 0 line. The haploid MA lines performed worse than the diploid MA lines. 

### Using lmer models to test for any differences between the ancestors

```{r}
null <- lmer(slope ~ 1 + (1|day), ctrl)
full <- lmer(slope ~ ploidy + (1|day), ctrl)
summary(full)

mod <- anova(null, full)
mod
```
Ploidy does not have a significant effect on fitness in the ancestors.


### Using lmer models with lineid and date as random effects.

##### Looking for MA-ploidy interaction:

```{r}
null <- lmer(slope ~ MA + ploidy + (1|day) + (1|lineid), df)
summary(null)
full <- lmer(slope ~ MA*ploidy + (1|day) + (1|lineid), df)
summary(full)

mod <- anova(null, full)
mod
```
No significant MA-ploidy interaction

##### Exploring effect of MA on the slope

```{r}
# lmer for effect of MA
null <- lmer(slope ~ 1 + ploidy + (1|day) + (1|lineid), df)
full <- lmer(slope ~ MA + ploidy + (1|day) + (1|lineid), df)
summary(full)

mod <- anova(null, full)
mod
```
The MA lines have different slopes than the ancestors

##### Exploring effect of ploidy on the slope 

```{r}
null <- lmer(slope ~ 1 + MA + (1|day) + (1|lineid), df)
full <- lmer(slope ~ ploidy + MA + (1|day) + (1|lineid), df)
summary(full)

mod <- anova(null, full)
mod
```
Haploids and diploids do not have different slopes

##### Combining information from the mutation rate dataset

```{r}
# read the files
mut <- read_delim("pombe_MA_data.txt")

# combine with the relative fitness data
trt2 <- trt %>%
  left_join(mut, by=c('label'='line'))
trt2 <- trt2 %>% select(ploidy.x, label, rel.fit, ploidy.y, ploidy_final,
                        n.SNM, n.indel)
trt2 <- trt2 %>% mutate(mutations = n.SNM + n.indel)
```

```{r, echo=FALSE}
trt2 %>% ggplot() +
  geom_point(aes(x=mutations , y=rel.fit, color=ploidy.x)) +
  geom_hline(yintercept = 0) +
  geom_point(data = trt2 %>% filter(label %in% aneuploids) %>%
               filter(ploidy.x == 'Diploid'),
             aes(x=mutations, y=rel.fit), color = 'red', shape='triangle', size = 3) +
  geom_point(data = trt2 %>% filter(label %in% aneuploids) %>%
               filter(ploidy.x == 'Haploid'),
             aes(x=mutations, y=rel.fit), color = 'blue', shape='triangle', size = 3) 
```

Some diploids have a relative fitness that is greater than 0. Diploids have more mutations than haploids.

##### Using a linear model to explore mutation rate and ploidy interaction

```{r}
mod <- lm(rel.fit ~ ploidy.x*mutations + ploidy.x + mutations, trt2)
summary(mod)
qqnorm(resid(mod))
```

##### Linear model with no mutation rate and ploidy interactions 

```{r}
mod <- lm(rel.fit ~ ploidy.x + mutations, trt2)
summary(mod)
qqnorm(resid(mod))
```

##### Linear model with only the mutations

```{r}
mod <- lm(rel.fit ~ mutations, trt2)
summary(mod)
qqnorm(resid(mod))

```

##### Combining the mutation dataset with the entire assay data

```{r}
mut_join <- trt2 %>% select(label, mutations)
df <- df %>% left_join(mut_join) %>% 
  mutate(mutations = ifelse(is.na(mutations), 0, mutations))

```

##### Mutation rate - ploidy interaction

```{r}
null <- lmer(slope ~ mutations + ploidy + (1|day) + (1|lineid), df)
summary(null)
full <- lmer(slope ~ mutations*ploidy + (1|day) + (1|lineid), df)
summary(full)

mod <- anova(null, full)
mod
```
No significant interaction in number of mutations and ploidy

##### Effect of mutations

```{r}
null <- lmer(slope ~ 1 + ploidy + (1|day) + (1|lineid), df)
full <- lmer(slope ~ mutations + ploidy + (1|day) + (1|lineid), df)
summary(full)

mod <- anova(null, full)
mod
```
No significant fitness effects of number of mutations.

###### Effect of ploidy

```{r}
null <- lmer(slope ~ 1 + mutations + (1|day) + (1|lineid), df)
full <- lmer(slope ~ ploidy + mutations + (1|day) + (1|lineid), df)
summary(full)

mod <- anova(null, full)
mod
```







```{r, include=FALSE}
# Looking at the predictions made by the lmer model (understanding the model)

# predictions <- predict(full, df[, c('MA', 'ploidy', 'day', 'lineid')])
# ggplot() +
#   geom_point(aes(x=df$slope, y=predictions)) +
#   geom_abline(slope = 1, intercept = 0, color='red') +
#   xlim(0, NA) +
#   ylim(0, NA)
# 
# ggplot() +
#   geom_point(aes(x=df$slope, y=fitted(full), color=df$day)) +
#   geom_abline(slope = 1, intercept = 0, color='red') +
#   xlim(0, NA) +
#   ylim(0, NA)
# 
# 
# ggplot() + geom_point(aes(x=fitted(full), y= resid(full),
#                           color=df$day)) #+ theme(legend.position = "none")
# 
# plot(full)
# summary(full)
# 
# ggplot() + geom_histogram(aes(x=resid(full)), fill='black')
# 
# qqnorm(resid(full))
```

